On Determinism of Game Engines used for Simulation-based Autonomous Vehicle Verification

### 2020-11-13 paper/cav_determinisim_main.tex
Added the setup guide. Done various little touches and corrections.

% The use of game engines for AV testing
Increasingly, the autonomous vehicle community is adopting game engines as simulation platforms to support the development and testing of vehicle control software. 
%
CARLA\footnote{\url{http://carla.org/}}, for instance, is an open-source simulator for autonomous driving that is implemented in the Unreal Engine\footnote{\url{https://www.unrealengine.com/}}, a real-time 3D creation environment for the gaming and film industry as well as other creative sectors~\cite{CARLA_paper}.

@@

\subsection{Definitions}
\subsubsection{Determinism}
% What is a deterministic system
Schumann describes determinism as the property of causality given a temporal development of events such that any state is completely determined by prior states~\cite{Schumann2010}. However, in the context of simulation this should be expanded to include not just prior states but also the history of actions taken by all actors. Therefore, a deterministic simulation will always produce the same result given the same history of prior states and actions.
% 
% Herein the term \textit{simulation variance} is used to refer to how simulation traces and outputs vary, when the same stimulus is provided repeatedly to the simulation environment under identical initial conditions. Specifically, we focus on the variance of actor paths within a given scenario.

@@

There are numerous game engines with their associated development environments which could be considered suitable for AV development, e.g. Unreal Engine\footnote{\url{https://www.unrealengine.com/}}, Unity\footnote{\url{https://unity.com/}}, CryEngine\footnote{\url{https://www.cryengine.com/}}. Specific autonomous driving research tools have been created to abstract and simplify the development environment, some of which are based on existing game engines, e.g.\ CARLA\footnote{\url{http://carla.org/}}, AirSim\footnote{\url{https://microsoft.github.io/AirSim/}}, Apollo\footnote{\url{http://apollo.auto/}}, and some have been developed for cloud based simulation, e.g.\ Nvidia Drive Constellation~\cite{nvidia_constellation}.
% \footnote{\url{https://www.nvidia.com/en-gb/self-driving-cars/drive-constellation/}}.

@@

% ======= Overview of a Game Loop
\subsection{Game Engine Overview} \label{GameLoopSection}

@@

The physics engine operates according to a time step, $dt$. %, because it makes everything simpler and more stable for physics and AI.
The shorter this time step is, the smoother the interpretation of the physical dynamics will be. %processing time it takes to catch up to game clock. % and the more deterministic the engine becomes and vice versa, as will be explained in section \ref{s:nondeterminisimSources}. 
%
% Fixing the physics time step to a constant $dt$ results in simpler and more stable physics simulation and AI. 
%
To use a fixed physics time step, the user's display refresh rate needs to be known in advance. This requires an update loop to take less than one render tick (one frame of real world time). Given the range of different hardware capabilities, a variable delta time is often implemented for game playing, taking the previous frame time as the next $dt$. However, variable $dt$ can lead to different outcomes in repeated tests and in some cases unrealistic physical representations~\cite{gaffer}. 
%
Semi-fixed or limited frame rates ensure $dt$ does not exceed some user-defined limit to meet a minimum standard of physical representation but allow computational headroom for slower hardware. Some engines provide sub-stepping which processes multiple physics calculations per frame at a greater CPU cost, e.g.\ Unreal Engine\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Physics/Substepping/}}. 
%
If the engine tries to render between physics updates, \textit{residual lag} can occur, which may result in frames arriving with a delay to the simulated physics.
%
Thus, extrapolation between frames may need to be performed to smooth transition between scenes. 
%
Note that both residual lag and extrapolation could affect perception stack testing.
% This is depicted in the lower part of Fig.~\ref{GameEngineLoopDiagram}. 
% \cite{GameEngineArchBook}\cite{GameProgPatternsBook}\todo[inline]{Should this still be cited? GC-No, it is cited elsewhere.}
%
% \todo[inline]{Why is this part of the figure? It would be clearer if it was in a separate figure that shows all the clocks involved and how they interact - depend on each other. GC-suggest we remove as the ticks are not relevant to the main argument or discussion.}
%
% Ideally the time step should be as short as possible, so that the simulation returns better accuracy of complex physical processes. %runs with high fidelity on fast machines. 
In exceptional cases where computational resources are scarce, the fixed time step can be greater than the time between render ticks %it takes to process an ``Update" inner loop on some (slow) machines, 
and the simulation will exhibit lag between input commands and rendered states, resulting in unsynchronised and unrealistic behaviour as can be experienced when games are executed on platforms not intended for gaming. 
% \todo[inline]{Explain how this lag may affect perception stack testing}

@@

However, if some aspects of the game loop are multi-threaded\footnote{\url{https://docs.unity3d.com/Manual/JobSystemMultithreading.html}}, then, even with a clear thread scheduling order, any background process may interrupt the otherwise deterministic sequence of events. This may, for example, alter the number of physics calculations that can be performed within the game loop and hence result in simulation variance.
%
Using multiple threads has been found to affect initialisation ordering for training machine learning models which can lead to unpredictable ordering of training data and non-deterministic behaviour~\cite{Sculley2015,Breck2017}.
%
\todo[inline,color=green!40]{GC addition: Using multiple threads has been found to affect initialisation ordering for training machine learning models which can lead to unpredictable ordering of training data and non-deterministic behaviour~\cite{Sculley2015,Breck2017}. }
%
Interference may also occur when a scheduler simply randomly selects from a set of threads with equal priority, resulting in variation of the thread execution order.

@@

Indeed, the developers of the de-bugging program \texttt{RR}\footnote{\url{https://github.com/mozilla/rr}} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory. This allowed control over the scheduling and execution order of threads promoting deterministic behaviour~\cite{acm-q-rr-interview}.
%
\todo[inline,color=green!40]{GC addition:Indeed, the developers of the de-bugging program \texttt{RR} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory. This allowed control over the scheduling and execution order of threads promoting deterministic behaviour~\cite{acm-q-rr-interview}. }

@@

\subsection{Game Engine Setup}
The type and version of the engine code executed should be considered, paying attention to the control of pseudo-random numbers, fixed physics calculation steps ($dt$), fixed actor navigation mesh, deterministic ego vehicle controllers and engine texture loading rates especially for perception stack testing. %
%
% It must be noted that if $dt$ cannot be fixed then deterministic results are unlikely.
% \todo[inline]{introduce actor navigation mesh GC-section on actor navigation brought forward}
% \todo[inline]{introduce ego vehicle earlier on GC added line 99}
For example, in Unreal Editor the \textit{unit}\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Performance/StatCommands/}} command can be used to monitor performance metrics such as \textit{Frame} which reports 
% to determine processing bottlenecks by indicating\todo[inline]{what does indicating mean? GC-displaying, returning or showing} 
the total time spent generating one frame, \textit{Game} for game loop execution time and \textit{Draw} for render thread time. 
%
With respect to perception stack testing, weather and lighting conditions in the game engine should be controlled as well as any other dynamic elements to the simulation environment, e.g.\ reflections from surface water, ensuring textures are not randomly generated.

\subsection{Actor Navigation}
There is existing evidence to suggest actor navigation, typically pedestrians, could be the cause of non-deterministic simulation behaviour~\cite{CARLABenchmark}.
% 
Unreal Engine is the underlying framework for the CARLA simulator which uses the A* algorithm for actor navigation\footnote{\url{https://www.oreilly.com/library/view/unreal-engine-4/9781784393120/ch04s03.html} from~\cite{newton2016unreal}}.
%
The A* algorithm~\cite{AStarBook} will give deterministic outcomes as long as the environment is deterministic~\cite{AirsimUnrealArticle, UnrealAIDocumentation}.

@@

\subsection{Actor Collisions}\label{S:Actor_Collisions}
Previous investigations into the Unreal Engine indicated that collisions between dynamic actors and solid objects, termed \textit{blocking physics bodies} in Unreal Engine documentation\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Physics/Collision/Overview/index.html}}, can lead to high simulation variance, \cite{TSLUnrealEngineTesting}. 
%
Collisions and the subsequent physics calculations that are processed, termed \textit{event hit} callback in Unreal Engine, were seen as potentially key aspects to the investigation into simulation variance.

@@

Within Unreal Engine there are numerous internal settings relating to the movement and interaction of physical bodies in the simulation. Settings can be adjusted to alter how actors interact and path plan via the navigation mesh of the environment, e.g. \textit{Contact Offset} and \textit{Navmesh Voxel Size}, or can be changed to improve the fidelity of physics calculations between game update steps, e.g. \textit{Physics Sub-Stepping} and \textit{Max Physics Delta Time}. Other options such as \textit{Enable Enhanced Determinism} were investigated along with running the engine from the command line with options for more deterministic behaviour \texttt{-deterministic}, floating point control \texttt{/fp:strict} and headless mode \texttt{-nullrhi} along with running the test as a packaged release by building and cooking.\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Deployment/Releasing/}} 
% \todo[inline]{Maybe we can introduce our initial study here? Picture of the man and moving bloack and a result showing the path traces in x,y etc?}
An initial study into the Unreal Engine using a pedestrian and a moving block was used to investigate simulation variance against these settings. The results were compared to a baseline of the default engine settings. 
%
However, none of these options improved simulation variance significantly and all internal setting were set restored to the default values. Details on this previous investigation can be found on the Trustworthy Systems GitHub~\cite{TSLUnrealEngineTesting}.

@@

\todo[inline,color=green!40]{GC Addition (para below):} 
%Issues around floating point precision and concurrency as previously mentioned notwithstanding, 
Practitioners should also be aware that many libraries for calculating variance itself may require attention to get precise results. For example the \texttt{numpy} method of variance is sensitive to the input precision and will return an incorrect answer if the wrong parameters are set~\cite{NumpyVar}. In \texttt{matlab}, the calculation of variance may switch from single thread to a multi-threaded execution not obviously apparent to the user when the input data size becomes large enough, opening up the potential for concurrency imprecision.\footnote{\url{https://uk.mathworks.com/matlabcentral/answers/317128-does-matlab-use-all-cores-by-default-when-running-a-program}}

@@

% ***************************************************
%  System Configuration and Screening
% ***************************************************
\subsection{System Configuration and Screening}\label{s:screening}
These tests were carried out on an Alienware Area 51 R5 with an i9 9960X processor with 64GB non-ECC DDR4 RAM at 2933MHz with an NVIDIA GeForce RTX 2080 GPU with 8GB GDDR6 RAM at 1515 MHz. Operating systems was Linux Ubuntu 18.04.4 LTS. To ensure reliability of results, each test was repeated 1000 times. Tests were carried out in CARLA (v0.9.6 and Unreal Engine v4.22) using synchronous mode with a fixed $dt$ of $0.05$s. To eliminate some of the potential sources of non-determinism outlined in Section~\ref{s:nondeterminisimSources} a series of screening tests and analyses were performed on our system. These were:

\begin{itemize}[leftmargin=*]
    \item System memory: \texttt{memtest86}\footnote{\url{https://www.memtest86.com/}} full suite of tests ran, all passed.
    \item Graphical memory: \texttt{cuda\_memtest}\footnote{see \url{https://github.com/ComputationalRadiationPhysics/cuda_memtest}} full details of tests found here~\cite{shi2009testing}, all tests passed.
    \item Non-uniform memory access: \texttt{numactl}\footnote{\url{https://linux.die.net/man/8/numactl}} was used to fix the simulator and test script to single cores but only a minor (2\%) improvement in simulation variance was observed and therefore not used for subsequent testing.
\end{itemize}

@@

\begin{figure*}[t]
    \centering
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionCG25_V4.pdf}
        \caption{}
        \label{CarsCollisionCG25}
    \end{subfigure}
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionPrePostV5.pdf}
        \caption{}
        \label{CarsCollisionPrePost}
    \end{subfigure}
    \caption{Vehicle to vehicle collision (test 2) showing (a) mean deviation against simulation time for 25\% resource utilisation and (b) mean deviation pre- and post-collision against resource utilisation. The simulation noise floor is shown in (a) which is the empirical lower limit of deviation for the hardware reported in this study.}
\end{figure*}

@@

\begin{figure*}[h]
    \centering
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollsionsCG25_V3.pdf}
        \caption{}
        \label{CarsPeopleCollsionsCG25}
    \end{subfigure}
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollisionPrePostV5.pdf}
        \caption{}
        \label{CarsPeopleCollisionPrePost}
    \end{subfigure}
    \caption{Vehicle to pedestrian collision (test 4) showing (a) maximum deviation against simulation time for 25\% stress and (b) maximum deviation pre- and post-collision for different resource utilisation levels.}
\end{figure*}

\begin{figure}[t]
    \centering
    \includegraphics[width=0.48\textwidth]{Other/Figures/NICE_analysis_summary_V3.pdf}
    \caption{Summary of investigating the effect of NICE priority setting for 0\% and 75\% additonal CPU \& GPU resource utilisation.}
    \label{NICEExperimentStressSummary}
\end{figure}

@@

\subsection{Process Scheduling Priority}
An investigation into the response of process scheduling on the simulation variance was undertaken. The experiment was repeated using test 1 but altering the process scheduling priority using the Ubuntu program \texttt{NICE}. Setting a higher priority for the simulator process with respect to the resource utilisation processes, it was possible to determine if scheduling could account for the increased simulation variance when the system is under stress.
%
To give a process a high priority a negative NICE value is set with the lowest being -20. To decrease the priority a positive value is set, up to +19. The default NICE value is 0.

The results presented in a box and whisker format, Fig.~\ref{NICEExperimentStressSummary}, show that decreasing the priority (right hand side of plot) has little effect on simulation variance when compared to a 0 NICE value (central plot bars). Increasing priority (left hand side of plot) reduced variance for the 75\% resource utilisation  but this did not account for all the difference in the observed results. 
%
This unaccounted difference in the variance may be due to the lack of  control that \texttt{NICE} has over the processes~\cite{Abanoub can you cite that quote you found here please}.

@@

\subsection{Simulator Settings}
The settings internal to the game engine or other simulation environment should be set to ensure a fixed physics time step, $dt$. If using CARLA a small fixed value, say $0.05s$, can be set by using the following: \texttt{setting.fixed\_delta\_seconds = 0.05}, whereas in Unity the default fixed time step is set to $0.02s$\footnote{\url{https://docs.unity3d.com/ScriptReference/MonoBehaviour.FixedUpdate.html}}. In CARLA, synchronous mode must be used to allow communication to external controllers which ensures no sensor data are passed out of order to the simulator which is particularly important if a complex ego controller is used~\cite{carla_sim_config}. 
%
The use of random numbers must be controlled through the use of fixed seeds which might be used to control variations of background effects, e.g. weather patterns, or the navigation of random pedestrian actors, external vehicle controllers or other clients connected to the simulation environment. 
%
Actors that navigate through the environment should use a fixed navigation mesh.

@@

% If the policy or operating system cannot be changed then s
Setting the simulator process to a higher priority may help to alleviate conflicting task scheduling which can be achieved by using, for example \texttt{TaskSettings.Priority} in Windows\footnote{\url{https://docs.microsoft.com/en-us/windows/win32/taskschd/tasksettings-priority}} or \texttt{NICE} in Linux\footnote{\url{https://linux.die.net/man/1/nice}}.

@@

% \subsection{Future Work}
A potential avenue of exploration is developing a version of the simulator which has been designed specifically for verification. In this version the simulator would be deterministic because the execution is suitably controlled along with sources of randomness and scheduling. A similar task was achieved with the \texttt{RR} program.\footnote{\url{https://github.com/mozilla/rr}}

@@

% ***************************************************
%  FUNDING, REFS
% ***************************************************
\section*{Acknowledgements}
This research has in part been funded by the ROBOPILOT and CAPRI projects. Both
projects are part-funded by the Centre for Connected and Autonomous
Vehicles (CCAV), delivered in partnership with Innovate UK under grant numbers
103703 (CAPRI) and 103288 (ROBOPILOT), respectively.
\balance
\printbibliography

>>>>

% The use of game engines for AV testing
Increasingly, the autonomous vehicle community is adopting game engines as simulation platforms to support the development and testing of vehicle control software. 
%
CARLA~\cite{carla_main_website}, for instance, is an open-source simulator for autonomous driving that is implemented in the Unreal Engine~\cite{UE4_main_website} a real-time 3D creation environment for the gaming and film industry as well as other creative sectors~\cite{CARLA_paper}.

@@

\subsection{Definitions (Refer to Fig.~\ref{variance_description} throughout this section)}
\subsubsection{Determinism}
% What is a deterministic system
Schumann describes determinism as the property of causality given a temporal development of events such that any state is completely determined by prior states~\cite{Schumann2010}. However, in the context of simulation this should be expanded to include not just prior states but also the history of actions taken by all actors. Therefore, a deterministic simulation will always produce the same result given the same history of prior states and actions.
% 
% Herein the term \textit{simulation variance} is used to refer to how simulation traces and outputs vary, when the same stimulus is provided repeatedly to the simulation environment under identical initial conditions. Specifically, we focus on the variance of actor paths within a given scenario.

@@

There are numerous game engines with their associated development environments which could be considered suitable for AV development, e.g. Unreal Engine~\cite{UE4_main_website}, Unity~\cite{Unity_main_website}, CryEngine~\cite{CryEngine_main_website}. Specific autonomous driving research tools have been created to abstract and simplify the development environment, some of which are based on existing game engines, e.g.\ CARLA~\cite{carla_main_website}, AirSim~\cite{AirSim_main_website}, Apollo~\cite{Apollo_main_website}, and some have been developed for cloud based simulation, e.g.\ Nvidia Drive Constellation~\cite{nvidia_constellation}.
% \footnote{\url{https://www.nvidia.com/en-gb/self-driving-cars/drive-constellation/}}.

@@

% ======= Overview of a Game Loop
% \subsection{Game Engine Overview} \label{GameLoopSection}

@@

The physics engine operates according to a time step, $dt$. %, because it makes everything simpler and more stable for physics and AI.
The shorter this time step is, the smoother the interpretation of the physical dynamics will be. %processing time it takes to catch up to game clock. % and the more deterministic the engine becomes and vice versa, as will be explained in section \ref{s:nondeterminisimSources}. 
%
% Fixing the physics time step to a constant $dt$ results in simpler and more stable physics simulation and AI. 
%
To use a fixed physics time step, the user's display refresh rate needs to be known in advance. This requires an update loop to take less than one render tick (one frame of real world time). Given the range of different hardware capabilities, a variable delta time is often implemented for game playing, taking the previous frame time as the next $dt$. However, variable $dt$ can lead to different outcomes in repeated tests and in some cases unrealistic physical representations~\cite{gaffer}. 
%
Semi-fixed or limited frame rates ensure $dt$ does not exceed some user-defined limit to meet a minimum standard of physical representation but allow computational headroom for slower hardware. Some engines provide sub-stepping which processes multiple physics calculations per frame at a greater CPU cost, e.g.\ Unreal Engine~\cite{UE4_substepping}. 
%
If the engine tries to render between physics updates, \textit{residual lag} can occur, which may result in frames arriving with a delay to the simulated physics.
%
Thus, extrapolation between frames may need to be performed to smooth transition between scenes. 
%
Note that both residual lag and extrapolation could affect perception stack testing.
% This is depicted in the lower part of Fig.~\ref{GameEngineLoopDiagram}. 
% \cite{GameEngineArchBook}\cite{GameProgPatternsBook}\todo[inline]{Should this still be cited? GC-No, it is cited elsewhere.}
%
% \todo[inline]{Why is this part of the figure? It would be clearer if it was in a separate figure that shows all the clocks involved and how they interact - depend on each other. GC-suggest we remove as the ticks are not relevant to the main argument or discussion.}
%
% Ideally the time step should be as short as possible, so that the simulation returns better accuracy of complex physical processes. %runs with high fidelity on fast machines. 
In exceptional cases where computational resources are scarce, the fixed time step can be greater than the time between render ticks %it takes to process an ``Update" inner loop on some (slow) machines, 
and the simulation will exhibit lag between input commands and rendered states, resulting in unsynchronised and unrealistic behaviour as can be experienced when games are executed on platforms not intended for gaming. 
% \todo[inline]{Explain how this lag may affect perception stack testing}

@@

However, if some aspects of the game loop are multi-threaded~\cite{unity_multithreading}, then, even with a clear thread scheduling order, any background process may interrupt the otherwise deterministic sequence of events. This may, for example, alter the number of physics calculations that can be performed within the game loop and hence result in simulation variance.
%
Using multiple threads has been found to affect initialisation ordering for training machine learning models which can lead to unpredictable ordering of training data and non-deterministic behaviour~\cite{Sculley2015,Breck2017}.
%
\todo[inline,color=green!40]{GC addition: Using multiple threads has been found to affect initialisation ordering for training machine learning models which can lead to unpredictable ordering of training data and non-deterministic behaviour~\cite{Sculley2015,Breck2017}. }
%
Interference may also occur when a scheduler simply randomly selects from a set of threads with equal priority, resulting in variation of the thread execution order.

@@

Indeed, the developers of the de-bugging program \texttt{RR}~\cite{RR_link} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory. This allowed control over the scheduling and execution order of threads promoting deterministic behaviour~\cite{acm-q-rr-interview}.
%
\todo[inline,color=green!40]{GC addition:Indeed, the developers of the de-bugging program \texttt{RR} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory. This allowed control over the scheduling and execution order of threads promoting deterministic behaviour~\cite{acm-q-rr-interview}. }

@@

\subsection{Game Engine Setup}
The type and version of the engine code executed should be considered, paying attention to the control of pseudo-random numbers, fixed physics calculation steps ($dt$), fixed actor navigation mesh, deterministic ego vehicle controllers and engine texture loading rates especially for perception stack testing. %
%
% It must be noted that if $dt$ cannot be fixed then deterministic results are unlikely.
% \todo[inline]{introduce actor navigation mesh GC-section on actor navigation brought forward}
% \todo[inline]{introduce ego vehicle earlier on GC added line 99}
For example, in Unreal Editor the \textit{unit}\cite{stat_commands} command can be used to monitor performance metrics such as \textit{Frame} which reports 
% to determine processing bottlenecks by indicating\todo[inline]{what does indicating mean? GC-displaying, returning or showing} 
the total time spent generating one frame, \textit{Game} for game loop execution time and \textit{Draw} for render thread time. 
%
With respect to perception stack testing, weather and lighting conditions in the game engine should be controlled as well as any other dynamic elements to the simulation environment, e.g.\ reflections from surface water, ensuring textures are not randomly generated.

\subsection{Actor Navigation}
There is existing evidence to suggest actor navigation, typically pedestrians, could be the cause of non-deterministic simulation behaviour~\cite{CARLABenchmark}.
% 
Unreal Engine is the underlying framework for the CARLA simulator which uses the A* algorithm for actor navigation~\cite{a_Star_oreilly}.
%
The A* algorithm~\cite{AStarBook} will give deterministic outcomes as long as the environment is deterministic~\cite{AirsimUnrealArticle, UnrealAIDocumentation}.

@@

\subsection{Actor Collisions}\label{S:Actor_Collisions}
Previous investigations into the Unreal Engine indicated that collisions between dynamic actors and solid objects, termed \textit{blocking physics bodies} in Unreal Engine documentation~\cite{collision_overview}, can lead to high simulation variance, \cite{TSLUnrealEngineTesting}. 
%
Collisions and the subsequent physics calculations that are processed, termed \textit{event hit} callback in Unreal Engine, were seen as potentially key aspects to the investigation into simulation variance.

@@

Within Unreal Engine there are numerous internal settings relating to the movement and interaction of physical bodies in the simulation. Settings can be adjusted to alter how actors interact and path plan via the navigation mesh of the environment, e.g. \textit{Contact Offset} and \textit{Navmesh Voxel Size}, or can be changed to improve the fidelity of physics calculations between game update steps, e.g. \textit{Physics Sub-Stepping} and \textit{Max Physics Delta Time}. Other options such as \textit{Enable Enhanced Determinism} were investigated along with running the engine from the command line with options for more deterministic behaviour \texttt{-deterministic}, floating point control \texttt{/fp:strict} and headless mode \texttt{-nullrhi} along with running the test as a packaged release by building and cooking~\cite{releasing_project}.
% \todo[inline]{Maybe we can introduce our initial study here? Picture of the man and moving bloack and a result showing the path traces in x,y etc?}
An initial study into the Unreal Engine using a pedestrian and a moving block was used to investigate simulation variance against these settings. The results were compared to a baseline of the default engine settings. 
%
However, none of these options improved simulation variance significantly and all internal setting were set restored to the default values. Details on this previous investigation can be found on the Trustworthy Systems GitHub~\cite{TSLUnrealEngineTesting}.

@@

\todo[inline,color=green!40]{GC Addition (para below):} 
%Issues around floating point precision and concurrency as previously mentioned notwithstanding, 
Practitioners should also be aware that many libraries for calculating variance itself may require attention to get precise results. For example the \texttt{numpy} method of variance is sensitive to the input precision and will return an incorrect answer if the wrong parameters are set~\cite{NumpyVar}. In \texttt{matlab}, the calculation of variance may switch from single thread to a multi-threaded execution not obviously apparent to the user when the input data size becomes large enough, opening up the potential for concurrency imprecision~\cite{matlab_parallel_computing}.

@@

% ***************************************************
%  System Configuration and Screening
% ***************************************************
\subsection{System Configuration and Screening}\label{s:screening}
These tests were carried out on an Alienware Area 51 R5 with an i9 9960X processor with 64GB non-ECC DDR4 RAM at 2933MHz with an NVIDIA GeForce RTX 2080 GPU with 8GB GDDR6 RAM at 1515 MHz. 
%
Operating systems was Linux Ubuntu 18.04.4 LTS. To ensure reliability of results, each test was repeated 1000 times. 
%
Tests were carried out in CARLA (v0.9.6 and Unreal Engine v4.22) using synchronous mode with a fixed $dt$ of $0.05$s. 
%
A detailed guide for reproducing these experiments along with the scripts used are provided on github\footnote{\url{https://github.com/TSL-UOB/CAV-Determinism/tree/master/CARLA_Tests_setup_guide}}.
%
To eliminate some of the potential sources of non-determinism outlined in Section~\ref{s:nondeterminisimSources} a series of screening tests and analyses were performed on our system. These were:

\begin{itemize}[leftmargin=*]
    \item System memory: \texttt{memtest86}~\cite{MemTest86} full suite of tests ran, all passed.
    \item Graphical memory: \texttt{cuda\_memtest}~\cite{cuda_memtest}\cite{shi2009testing}.
    \item Non-uniform memory access: \texttt{numactl}~\cite{numactl_NUMA} was used to fix the simulator and test script to single cores but only a minor (2\%) improvement in simulation variance was observed and therefore not used for subsequent testing.
\end{itemize}

@@

\begin{figure*}[t]
    \centering
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionCG25_V4.pdf}
        \caption{}
        \label{CarsCollisionCG25}
    \end{subfigure}
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionPrePostV5.pdf}
        \caption{}
        \label{CarsCollisionPrePost}
    \end{subfigure}
    \caption{Vehicle to vehicle collision (test 2) showing (a) maximum deviation against simulation time for 25\% resource utilisation and (b) maximum deviation pre- and post-collision against resource utilisation. The simulation noise floor is shown in (a) which is the empirical lower limit of deviation for the hardware reported in this study.}
\end{figure*}

@@

\begin{figure*}[h]
    \centering
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollsionsCG25_V3.pdf}
        \caption{}
        \label{CarsPeopleCollsionsCG25}
    \end{subfigure}
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollisionPrePostV5.pdf}
        \caption{}
        \label{CarsPeopleCollisionPrePost}
    \end{subfigure}
    \caption{Vehicle to pedestrian collision (test 4) showing (a) maximum deviation against simulation time for 25\% resource utilisation and (b) maximum deviation pre- and post-collision for different resource utilisation levels.}
\end{figure*}

\begin{figure}[t]
    \centering
    \includegraphics[width=0.48\textwidth]{Other/Figures/NICE_analysis_summary_V4.pdf}
    \caption{Summary of investigating the effect of NICE priority setting for 0\% and 75\% additonal CPU \& GPU resource utilisation.}
    \label{NICEExperimentStressSummary}
\end{figure}

@@

\subsection{Process Scheduling Priority}
An investigation into the response of process scheduling on the simulation variance was undertaken. The experiment was repeated using test 1 but altering the process scheduling priority using the Ubuntu program \texttt{NICE}. Setting a higher priority for the simulator process with respect to the resource utilisation processes, it was possible to determine if scheduling could account for the increased simulation variance when the system is under high resource utilisation.
%
To give a process a high priority a negative NICE value is set with the lowest being -20. To decrease the priority a positive value is set, up to +19. The default NICE value is 0.

The results presented in a box and whisker format, Fig.~\ref{NICEExperimentStressSummary}, show that decreasing the priority (right hand side of plot) has little effect on simulation variance when compared to a 0 NICE value (central plot bars). Increasing priority (left hand side of plot) reduced variance for the 75\% resource utilisation but this did not account for all the difference in the observed results. 
%
This unaccounted difference in the variance may be due to the lack of  control that \texttt{NICE} has over the processes\footnote{\url{https://askubuntu.com/questions/656771/process-niceness-vs-priority}}.
%~\cite{Abanoub can you cite that quote you found here please}.

@@

\subsection{Simulator Settings}
The settings internal to the game engine or other simulation environment should be set to ensure a fixed physics time step, $dt$. If using CARLA a small fixed value, say $0.05s$, can be set by using the following: \texttt{setting.fixed\_delta\_seconds = 0.05}, whereas in Unity the default fixed time step is set to $0.02s$~\cite{MonoBehaviour_unity}. In CARLA, synchronous mode must be used to allow communication to external controllers which ensures no sensor data are passed out of order to the simulator which is particularly important if a complex ego controller is used~\cite{carla_sim_config}. 
%
The use of random numbers must be controlled through the use of fixed seeds which might be used to control variations of background effects, e.g. weather patterns, or the navigation of random pedestrian actors, external vehicle controllers or other clients connected to the simulation environment. 
%
Actors that navigate through the environment should use a fixed navigation mesh.

@@

% If the policy or operating system cannot be changed then s
Setting the simulator process to a higher priority may help to alleviate conflicting task scheduling which can be achieved by using, for example \texttt{TaskSettings.Priority} in Windows~\cite{TaskSettingWindows} or \texttt{NICE} in Linux~\cite{Nice_linux}.

@@

% \subsection{Future Work}
A potential avenue of exploration is developing a version of the simulator which has been designed specifically for verification. In this version the simulator would be deterministic because the execution is suitably controlled along with sources of randomness and scheduling. A similar task was achieved with the \texttt{RR} program~\cite{RR_link}.

@@

% ***************************************************
%  FUNDING, REFS
% ***************************************************
\section*{Acknowledgements}
This research has in part been funded by the ROBOPILOT and CAPRI projects. Both
projects are part-funded by the Centre for Connected and Autonomous
Vehicles (CCAV), delivered in partnership with Innovate UK under grant numbers
103703 (CAPRI) and 103288 (ROBOPILOT), respectively.
\printbibliography
